The Functional Therapeutic Chemical Classification System

### 2013-10-15 reply-to-editors.txt

Dear Editors,

We read with care the comments provided by the reviewers. We hereby present a modified version of the manuscript, based on their valuable input. Reviewer 1 asked for a more detailed presentation of the reasoning algorithm and methodology. In this regard, we have provided a supplementary reference to an introductory text. It is challenging to add more information to the main document, as reasoning services and tools have been developed by other research groups and are simply used for standard tasks in the context of our work. We rely on the W3C OWL2 international standard to interact with reasoning engines, for which extensive documentation is available. Moreover, we specially wrote an introductory text to description logics, available on the wiki, for biomedical users non familiar with the domain. Ideally description logics would be presented in greater details in the main manuscript, unfortunately and because of space concerns we decided to focus on the presentation of the biomedical novelty brought by our project. Understanding in details the methodology requires some knowledge about description logics (as noticed by Reviewer 2 too), yet we believe that Figure 1 provides a clear simplified overview. Reviewer 1 did not manage to find the Gene Ontology (GO) annotation "positive regulation of blood coagulation" for the protein Prothrombin (P00734). The annotation exists and is present on the GO annotations website: http://www.ebi.ac.uk/QuickGO/GProtein?ac=P00734. Reviewer 1 understood correctly the methodology, yet probably missed the link "Complete GO annotation..." on the Uniprot website, leading to the complete list where the mentioned annotation can be found. Reviewer 1 also asks for "how much of the different results between ATC and FTC". The relation between the two resources is described in section 2.1 "Evaluation" and furthermore on the website of the application (https://www.ebi.ac.uk/chembl/ftc/evaluation/), where we compare equivalent classes and derive metrics (precision and recall) summarising the overlap between these resources. The section 3.2 also discusses the interpretation of the evaluation and the differences/similarity between FTC/ATC. It is unfortunately not clear to us what are the specific requirements of reviewer 1 regarding this revision.
Reviewer 2 asked us to illustrate the potential of the resource with a known repurposing example. We chose the thalidomide to fulfil this legitimate requirement, as it is possibly one of the best known repositioning case. We have added a paragraph to the manuscript (end of section 2.2.1) explaining how the transgenic effect of the molecule is described in the FTC and how the new indications of the drug are captured (anti-cancer and immunomodulatory activities, nowadays routinely prescribed to patients). We briefly discussed the relationship between our resource and the work done in Bioinformatics 28.16 (2012): 2169-2175 and the Open PHACTS platform. Moreover, the methodology section further references an omitted similar approach presented in Nucleic Acids Res. 2011 Oct;39(18):e119. Reviewer 2 pointed the confusing number of drugs and their origin. This section was clarified in the manuscript (section 2); we used the FDA-approved drugs as described in DrugBank and interacting with at least one human protein target. Reviewer 2 asked whether the outliers points present in Figure 4 could have different interpretation; the text was extended accordingly, and the meaning of the data points summarily discussed. Reviewer 2 expressed concerns regarding the user interaction with the resource, as a large number of new classes are present. We argue that the web application greatly improves the accessibility of the FTC's content. Each FTC category is described in details on the web application, and is further linked to related terms for the user to browse. Moreover, a text search functionality is available (https://www.ebi.ac.uk/chembl/ftc/search/), in order to easily find categories or drugs of interest. More search functionality might be added in the future, depending on the adoption of the classification.

### 2013-10-15 document.tex
ready for resubmission

\section{Motivation:}
Drug repurposing is the discovery of new indications for compounds that have already been approved and used in a clinical setting. 
Recently, some computational approaches have been suggested to unveil new opportunities in a systematic fashion, by taking 
into consideration gene expression signatures or chemical features for instance. 
We present here a novel method based on knowledge integration using semantic technologies, 
to capture the functional role of approved chemical compounds.
\section{Results:}
In order to computationally generate repositioning hypotheses, we used the Web Ontology Language (OWL) to 
formally define the semantics of over 20'000 terms with axioms to correctly denote various modes of action (MoA). Based on an 
integration of public data, we have automatically assigned over a thousand of approved drugs into 
these MoA categories. The resulting new resource is called the Functional Therapeutic Chemical Classification System (FTC) 
and was further evaluated against the content of the traditional Anatomical Therapeutic Chemical Classification System (ATC). 
We illustrate how the new classification can be used to generate drug repurposing hypotheses, using Alzheimer’s disease as a use-case.

@@

\begin{figure*}[!tpb]%figure1
\centerline{\includegraphics{fig1.png}}
\caption{The diagram gives an overview of the integrated resources. (A) The name of FTC categories representing MoAs are derived from the GO terms representing the molecular 
functions and biological processes. A 
prefix \emph{anti} or \emph{pro} is appended to the original 
GO term, as well as the word \emph{agent} as suffix. (B) Each of the new FTC class has a logical
equivalent definition assigned to it (axiom), representing the necessary and sufficient conditions for a drug to be classified
in the corresponding MoA class.
(C) The content of various databases is incorporated and linked using the FTC specific logical properties. (D) Finally a reasoner classifies the knowledge
base and assigns drugs to MoA classes based on whether or not a definition can be satisfied. For example, the drug
\emph{ximelagatran} will be assigned as member of the category \emph{Anti-blood coagulation agent} because of the logical
links \emph{ximelagatran negatively-perturbs prothrombin} and \emph{prothrombin involved-in positive regulation of blood coagulation}.
The taxonomic structure of the
FTC appears also in the reasoning step, from the entailment of the equivalent definitions.}\label{fig:01}
\end{figure*}

Drug repurposing is the use of known active compounds for new therapeutic indications \citep{Sanseau01072011}. 
When administered in a living organism, a compound can indeed play 
various roles and affect different biological processes (called \emph{mode of action}, MoA); 
accurately identifying these different functions 
helps to predict the potential side-effects a drug could have and can also lead to 
interesting repurposing opportunities \citep{Medina-Franco2013}. For instance, \emph{sildenafil} was initially developed 
to relieve angina pectoris symptoms and has been repurposed towards erectile dysfunction during the clinical trials \citep{Ashburn2004} when a new function of the target enzyme was discovered. Approved compounds are 
attractive because they have been extensively studied and have by definition already successfully passed clinical trials, where 
most drugs fail because of safety or efficacy issues.
There is an increasing number of approaches to predict repurposing opportunities using computational methods (see \cite{dudley2011exploiting} or \cite{andronis2011literature} for recent reviews). Most 
methods operate on the profiles of physicochemical descriptors derived from molecular structures \citep{Haupt2011}. Other methods characterise 
the drugs on more abstract levels, such as the gene expression signature \citep{Iorio2010} or 
via the reported side-effects \citep{Campillos2008}. 
These approaches 
have in common to look for similarities within existing drugs and forward similar compounds as repurposing hypotheses.
A feature of particular interest to describe drugs is the MoA. According to Wikipedia, 
the MoA describes \emph{a functional or anatomical change, 
at the cellular level, resulting from the exposure of a living organism to a substance}. For instance 
terms such as \emph{transcriptional regulation agent} or \emph{anticoagulant} define MoAs and characterise the 
roles of a certain type of drugs. The 
MoA abstracts over the relations between molecular functions, protein targets and drug activities; it is the central concept linking a chemical structure to a set of biological activities.
Intuitively, the indication of a drug logically depends on its MoA.
Despite its widespread use in drug discovery, the MoA has not been used yet as a descriptor for repurposing analyses. One reason for this 
might be the challenge of formally defining MoAs. Indeed, MoAs are terms or categories, it is not possible to represent 
them straightforwardly with values and numbers like one can do for a 3D molecular structure or for a gene expression profile. 
Nonetheless, the meaning of a concept can be formalised with controlled vocabularies and ontologies \citep{Gruber1995}; 
originating from description logics, such frameworks help to formalise the semantics of symbols and strings of characters with explicit axioms. 
In an ontology or knowledge base, 
\emph{concepts} (interchangeable with \emph{category}, \emph{term} and \emph{class} in this manuscript) are 
organised and linked following the logical type of relation they have among them. 
In the Gene Ontology (GO) for example \citep{Ashburner2000}, biological processes and molecular functions terms are manually curated  and their meaning specified by the relation types linking two GO terms.
MoA definitions are present in other classifications such as the Medical Subject Headings \citep{Nelson2004} or the Chemical Entities of 
Biological Interest \citep{Hastings2012} for example. The Anatomical Therapeutic 
Chemical Classification System (ATC) \citep{world2000anatomical} also describes to some 
extent the action of drugs at the anatomical level. All these resources are valuable for the community as a source of carefully 
manually curated information. Moreover, the categories described in these classification systems are sometimes used to annotate 
drugs: For instance the compound \emph{sildenafil} has been manually annotated as \emph{vasodilator agent} (CHEBI:35620 or MeSH:D27.505.954.411.918).
The classifications mentioned previously are not specially designed for drug repurposing; they purposefully report only the 
well-known and major MoAs of chemical compounds. The pharmacological spectrum of each drug is not necessarily well covered, yet 
it would be the best way to predict new indications. In our context, an ideal knowledge base would feature the known MoAs of a 
drug as well as some predicted ones to be tested in experiments. The MoA categories should derive and scale over primary molecular 
evidences exposed in biomedical databases, in an automated way.
To address the lack of systematic MoA annotations, we have implemented the Functional Therapeutic Chemical 
Classification System (FTC), presented here in this manuscript. The FTC is automatically built by leveraging the content 
of various biomedical databases using description logics and automated reasoning. Over 20'000 new MoA categories are defined 
in the resource and further populated with approved drugs using the Web Ontology Language (OWL) in combination with a reasoner.
The population step takes in account the type of pharmacological action, the molecular targets of the drug and their involvement
into biological processes. Drugs can exhibit several MoAs, and the same MoA can be reached through different mechanisms.
Most of the drugs are present in multiple FTC categories, reflecting the various roles a compound can play inside a biological system which can serve as starting point for repurposing. 
The resource was evaluated against the ATC, traditional classification scheme introduced before. We present as well some preliminary 
analyses over the data, by looking at the relation between the MoA and the indication of a compound using semantic similarity. 
Finally, we illustrate also how the FTC can be used as a pharmacological toolbox to analyse drug repurposing for Alzheimer’s disease.

@@

The knowledge base behind the FTC is built by integrating information coming from various sources. 
The GO terms serve as template to create the FTC categories describing the MoAs; DrugBank \citep{Knox2011} provides the 
known links between drugs and their protein targets and Uniprot \citep{TheUniprotConsortium2013} maps targets to their 
respective GO annotations \citep{Dimmer2012}. 
Drugs are further assigned into MoA categories according to the OWL constructs and axioms defined in the 
FTC. A reasoner, a program capable of understanding such axioms, performed this task (see \emph{Availability and Implementation} for details). 
The process to build the FTC is summarised in Figure~\ref{fig:01} alongside an example.
The core step is the generation of axiomatic representations of MoAs by decomposing GO types into positive and negative regulations of biomolecular functions and processes. 
The help of reasoning techniques we can further derive and assign MoA across the knowledge base to given drugs. It requires a few seconds (four processing cores, 4 GB RAM) to classify the knowledge base (ELK reasoner, \citep{Kazakov2011}). 
Other OWL reasoners (e.g, Hermit, Pellet, etc.) disqualified mainly due to long processing time (data not shown). 
The FTC forms a taxonomic structure as illustrated 
on Figure~\ref{fig:02}, which arises when the reasoner classifies the knowledge base. In general, categories may have multiple parents and multiple children (see {{https://www.ebi.ac.uk/chembl/ftc}} for interactive use).
In total there are 1'280 FDA-approved DrugBank compounds (chemical and biotherapeutics) associated with 1'264 human protein targets, where each drug is acting on at least one human protein target.
The FTC introduces 23'353 new categories new categories describing the mode and mechanism of action of therapeutic compounds. 
4'289 of these categories belong to the biological processes in GO and 19'064 to the molecular functions. A summary of the metrics behind the latest 
build is available online at {{https://www.ebi.ac.uk/chembl/ftc/evaluation/}}.
Out of all FTC categories, 1'432 
categories (\textgreater 6\%) directly contain at least one approved drug. This number increases up to 2'532 (\textgreater 11\%) 
when direct and 
indirect drugs are considered. FTC categories not containing drugs (e.g, FTC\_A0001771 - \emph{Anti-immunological synapse 
formation agent}) represent MoAs for which no approved compounds has qualified yet or that have not been identified  
as such in the FTC.

@@

\subsection{Evaluation}
The content of the FTC has been evaluated against the drug categorisation of the ATC, which has been produced by manual curation and serves as a gold standard. 
\emph{A priori}, both resources serve different purposes and as a consequence, the evaluation has to take this into consideration (cf section \ref{interpreting}). 
The full methodology behind the evaluation is described in the section 6 of the supplementary material (\emph{Evaluation methodology}).
Briefly, for 68 categories from the FTC we can manually identify a set of semantically equivalent categories in the ATC. We call these equivalent categories the \emph{evaluation points} (see supplementary material section 6.1 for details). 
All drugs from each evaluation point were then assessed to determine the quality of the FTC against our gold standard, i.e.\ the ATC.
For example, the FTC category \emph{Anti-hydrogen:potassium-exchanging ATPase activity agent} (FTC\_A0008900) has 
been manually asserted as equivalent to the ATC category \emph{Proton pump inhibitors} (A02BC). A summary of this 
evaluation point is furthermore available online at {{https://www.ebi.ac.uk/chembl/ftc/evaluation/FTC\_A0008900}}.
For 1'280 DrugBank compounds in the FTC, 1'134 are also present in the ATC, therefore only those were 
considered. The \emph{evaluation points} cover a total of 471 DrugBank compounds or around 41\% of common drugs to both classifications.
Out of these, 275 compounds are true positives, i.e.\ they match both, the FTC and ATC categories for a given evaluation point.
The \emph{proton pump inhibitor} 
evaluation point is such a case where all the drugs (\emph{omeprazole, esomeprazole, pantoprazole, lansoprazole, rabeprazole}) are 
present both in the FTC category and in the corresponding ATC categori(es). The total number of compounds from an ATC category but 
where we could not identify a corresponding FTC category is 35 (false negatives). Finally, 280 compounds are present in a FTC class but not in any 
corresponding ATC category (false positives). Overall we derive a 
recall of 89\%; this percentage indicates that the automatic build of the FTC covers a good portion of the content 
already present in the ATC. The precision of 50\% shows that the FTC contains for a given MoA many more drugs than the 
equivalent ATC categories. This result was expected and comes the original idea behind the FTC: Representing in a systematic fashion
the implicit and explicit MoAs of drugs, in particular the ones not already indexed by current classification scheme.

@@

\subsubsection{Polypharmacology spectrum}
The more information on a drug’s molecular targets and their physiological roles, the more opportunities 
exist to re-orient a drug into doing something new. The therapeutic agents described in the FTC can have several MoAs, 
i.e.\ may be acting on different biomolecular functions or processes, 
which demonstrates the intrinsic polypharmacology of the approved compounds. Figure \ref{fig:03} illustrates the polypharmacology 
spectrum by showing the distribution of number of MoAs per compound. When only direct categories are considered, compounds belong on average
to 13.5 MoA categories. This number increases to 61.2 when parent categories are taken into consideration (super classes).
Not all the MoAs are relevant to a disease, 
some FTC categories are particularly abstract (e.g, \emph{Anti-biological process agent}) yet they represent discrete 
categories to which the drug belongs with an explicit and clear meaning. These discrete MoAs are a good starting 
point to understand what a compound can do when administered in a human system.
Compound's polypharmacology is 
well represented in the FTC, as shown by the numerous MoAs each approved drug can exhibit.
We decided to further look at a well-known repositioning example, 
in order to see whether the FTC was suitable to identify the new
uses of an old drug. We picked the \emph{thalidomide} for 
this exercise ({{https://www.ebi.ac.uk/chembl/ftc/agent/DB01041}}). The molecule was 
first indicated to treat morning sickness
in pregnant women, but has been quickly abandoned after its developmental toxicity has been discovered in newborns. 
The accepted molecular mechanism behind the side-effect is  
an impairment of the angiogenic process responsible for the development of members, affecting in particular the limbs \citep{therapontos2009thalidomide}.
We found that the \emph{thalidomide} was accurately classified as \emph{Anti-cell migration involved in sprouting angiogenesis agent} in the FTC, capturing the known toxicity of the drug. Furthermore, the \emph{thalidomide} is currently investigated for a multitude of new usages, in particular for anti-cancer and immunomodulatory activities among others \citep{teo2005thalidomide}.
These new indications are well represented in the FTC too, for 
example by the categories \emph{Anti-vascular endothelial growth factor production agent} or 
\emph{Anti-cell division agent} for neoplastic activities, 
or by the classes \emph{Anti-cytokine secretion agent} and \emph{Anti-I-kappaB kinase/NF-kappaB cascade agent} for its effect
on the immune system. These observations demonstrate that the FTC can successfully capture the 
molecular reasons behind the repositioning of an old compound. 
Moreover the classification can also provide valuable insight regarding
potential toxicity too.

\begin{figure*}[!tpb]%figure4
\centerline{\includegraphics{fig4.png}}
\caption{Pairwise comparison of MoAs similarities. Therapeutic indications are represented by ATC categories which are the colours on the side.
For instance, the compound reteplase (DB00015) has the ATC code B01AD07, 
which appears as \emph{B} (dark orange) on the plot.
Only the first ATC level 
is considered. The reader can refer to Supplementary Figure 1A for a two ATC level granularity. 
The similarity descriptor ranges from 0 (not similar - white) to 1 (identical - black).
Some compounds belong to multiple ATC categories (\emph{Multiple}) and some others do not have
an ATC code (\emph{NoCategory}).
The average similarity of drugs present in the same therapeutic category 
is significantly higher on average when separately compared to all other indications.}\label{fig:04}
\end{figure*}

>>>>

\section{Motivation:}
Drug repositioning is the discovery of new indications for compounds that have already been approved and used in a clinical setting. 
Recently, some computational approaches have been suggested to unveil new opportunities in a systematic fashion, by taking 
into consideration gene expression signatures or chemical features for instance. 
We present here a novel method based on knowledge integration using semantic technologies, 
to capture the functional role of approved chemical compounds.
\section{Results:}
In order to computationally generate repositioning hypotheses, we used the Web Ontology Language (OWL) to 
formally define the semantics of over 20'000 terms with axioms to correctly denote various modes of action (MoA). Based on an 
integration of public data, we have automatically assigned over a thousand of approved drugs into 
these MoA categories. The resulting new resource is called the Functional Therapeutic Chemical Classification System (FTC) 
and was further evaluated against the content of the traditional Anatomical Therapeutic Chemical Classification System (ATC). 
We illustrate how the new classification can be used to generate drug repurposing hypotheses, using Alzheimer’s disease as a use-case.

@@

\begin{figure*}[!tpb]%figure1
\centerline{\includegraphics{fig1.png}}
\caption{The diagram gives an overview of the integrated resources. (A) The name of FTC categories representing MoAs 
are directly derived from the GO terms representing the molecular 
functions and biological processes. A 
prefix \emph{anti} or \emph{pro} is appended to the original 
GO term, as well as the word \emph{agent} as suffix (e.g, the GO term \emph{blood coagulation} serves as a template to create the
\emph{Anti-blood coagulation agent} and \emph{Pro-blood coagulation agent} categories). (B) Each of the new FTC class has a logical
equivalent definition assigned to it (axiom), representing the necessary and sufficient conditions for a drug to be classified
in the corresponding MoA class.
(C) The content of various databases is incorporated and linked using the FTC specific logical properties. (D) Finally a reasoner classifies the knowledge
base and assigns drugs to MoA classes based on whether or not a definition can be satisfied. For example, the drug
\emph{ximelagatran} will be assigned as member of the category \emph{Anti-blood coagulation agent} because of the logical
links \emph{ximelagatran negatively-perturbs prothrombin} and \emph{prothrombin involved-in positive regulation of blood coagulation}.
The taxonomic structure of the
FTC appears also in the reasoning step, from the entailment of the equivalent definitions.}\label{fig:01}
\end{figure*}

Drug repurposing is the use of known active compounds for new therapeutic indications \citep{Sanseau01072011}. 
When administered in a living organism, a compound can indeed play 
various roles and affect different biological processes (called \emph{mode of action}, MoA); 
accurately identifying these different functions 
helps to predict the potential side-effects a drug could have and can also lead to 
interesting repurposing opportunities \citep{Medina-Franco2013}. For instance, \emph{sildenafil} was initially developed 
to relieve angina pectoris symptoms and has been repurposed towards erectile dysfunction during the clinical trials \citep{Ashburn2004} when a new function of the target enzyme was discovered. Approved compounds are 
attractive because they have been extensively studied and have by definition already successfully passed clinical trials, where 
most drugs fail because of safety or efficacy issues.
There is an increasing number of approaches to predict repurposing opportunities using computational methods (see \cite{dudley2011exploiting} or \cite{andronis2011literature} for recent reviews). Most 
methods operate on the profiles of physicochemical descriptors derived from molecular structures \citep{Haupt2011}. Other methods characterise 
the drugs on more abstract levels, such as the gene expression signature \citep{Iorio2010} or 
via the reported side-effects \citep{Campillos2008}. 
These approaches 
have in common to look for similarities within existing drugs and forward similar compounds as repurposing hypotheses.

A feature of particular interest to describe drugs is the MoA. According to Wikipedia, 
the MoA describes \emph{a functional or anatomical change, 
at the cellular level, resulting from the exposure of a living organism to a substance}. For instance 
terms such as \emph{transcriptional regulation agent} or \emph{anticoagulant} define MoAs and characterise the 
roles of a certain type of drugs. The 
MoA abstracts over the relations between molecular functions, protein targets and drug activities; it is the central concept linking a chemical structure to a set of biological activities.
Intuitively, the indication of a drug logically depends on its MoA.
Despite its widespread use in drug discovery, the MoA has not been used yet as a descriptor for repurposing analyses. One reason for this 
might be the challenge of formally defining MoAs. Indeed, MoAs are terms or categories, it is not possible to represent 
them straightforwardly with values and numbers like one can do for a 3D molecular structure or for a gene expression profile. 
Nonetheless, the meaning of a concept can be formalised with controlled vocabularies and ontologies \citep{Gruber1995}; 
originating from description logics, such frameworks help to formalise the semantics of symbols and strings of characters with explicit axioms. 
In an ontology or knowledge base, 
\emph{concepts} (interchangeable with \emph{category}, \emph{term} and \emph{class} in this manuscript) are 
organised and linked following the logical type of relation they have among them. 
In the Gene Ontology (GO) for example \citep{Ashburner2000}, biological processes and molecular functions terms are manually curated  and their meaning specified by the relation types linking two GO terms.
MoA definitions are present in other classifications such as the Medical Subject Headings \citep{Nelson2004} or the Chemical Entities of 
Biological Interest \citep{Hastings2012} for example. The Anatomical Therapeutic 
Chemical Classification System (ATC) \citep{world2000anatomical} also describes to some 
extent the action of drugs at the anatomical level. All these resources are valuable for the community as a source of carefully 
manually curated information. Moreover, the categories described in these classification systems are sometimes used to annotate 
drugs: For instance the compound \emph{sildenafil} has been manually annotated as \emph{vasodilator agent} (CHEBI:35620 or MeSH:D27.505.954.411.918).
The classifications mentioned previously are not specially designed for drug repurposing; they purposefully report only the 
well-known and major MoAs of chemical compounds. The pharmacological spectrum of each drug is not necessarily well covered, yet 
it would be the best way to predict new indications.

In our context, an ideal knowledge base would feature the known MoAs of a 
drug as well as some predicted ones to be tested in experiments. The MoA categories should derive and scale over primary molecular 
evidences exposed in biomedical databases, in an automated way.
To address the lack of systematic MoA annotations, we have implemented the Functional Therapeutic Chemical 
Classification System (FTC), presented here in this manuscript. The FTC is automatically built by leveraging the content 
of various biomedical databases using description logics and automated reasoning. Over 20'000 new MoA categories are defined 
in the resource and further populated with approved drugs using the Web Ontology Language (OWL) in combination with a reasoner.
The population step takes in account the type of pharmacological action, the molecular targets of the drugs and their involvement
into biological processes. Drugs can exhibit several MoAs, and the same MoA can be reached through different mechanisms.
Most of the drugs are present in multiple FTC categories, reflecting the various roles a compound can play inside a biological system which can serve as starting point for repurposing. 
The resource was evaluated against the ATC, traditional classification scheme introduced before. We present as well some preliminary 
analyses over the data, by looking at the relation between the MoA and the indication of a compound using semantic similarity. 
Finally, we illustrate also how the FTC can be used as a pharmacological toolbox to analyse drug repurposing for Alzheimer’s disease.

@@

The knowledge base behind the FTC is built by integrating information coming from various sources. 
The GO terms serve as template to create the FTC categories describing the MoAs; DrugBank \citep{Knox2011} provides the 
known links between drugs and their protein targets and Uniprot \citep{TheUniprotConsortium2013} maps targets to their 
respective GO annotations \citep{Dimmer2012}. 
Drugs are further assigned into MoA categories according to the OWL constructs and axioms defined in the 
FTC. A reasoner, a program capable of understanding such axioms, performed this task (see \emph{Availability and Implementation} for details). 
The process to build the FTC is summarised in Figure~\ref{fig:01} alongside an example.
The core step is the generation of axiomatic representations of MoAs by decomposing GO types into positive and negative regulations of biomolecular functions and processes. 
The help of reasoning techniques we can further derive and assign MoA across the knowledge base to given drugs. It requires a few seconds (four processing cores, 4 GB RAM) to classify the knowledge base (ELK reasoner, \citep{Kazakov2011}). 
Other OWL reasoners (e.g, Hermit, Pellet, etc.) disqualified mainly due to long processing time (data not shown).

The FTC forms a taxonomic structure as illustrated 
on Figure~\ref{fig:02}, which arises when the reasoner classifies the knowledge base. In general, categories may have multiple parents and multiple children (see {{https://www.ebi.ac.uk/chembl/ftc}} for interactive use).
In total there are 1'280 FDA-approved DrugBank compounds (chemical and biotherapeutics) associated with 1'264 human protein targets, where each drug is acting on at least one human protein target.
The FTC introduces 23'353 new categories new categories describing the mode and mechanism of action of therapeutic compounds. 
4'289 of these categories belong to the biological processes in GO and 19'064 to the molecular functions. A summary of the metrics behind the latest 
build is available online at {{https://www.ebi.ac.uk/chembl/ftc/evaluation/}}.
Out of all FTC categories, 1'432 
categories (\textgreater 6\%) directly contain at least one approved drug. This number increases up to 2'532 (\textgreater 11\%) 
when direct and 
indirect drugs are considered. FTC categories not containing drugs (e.g, FTC\_A0001771 - \emph{Anti-immunological synapse 
formation agent}) represent MoAs for which no approved compounds has qualified yet or that have not been identified  
as such in the FTC.

@@

\subsection{Evaluation}
The content of the FTC has been evaluated against the drug categorisation of the ATC, which has been produced by manual curation and serves as a gold standard. 
\emph{A priori}, both resources serve different purposes and as a consequence, the evaluation has to take this into consideration (cf section \ref{interpreting}). 
The full methodology behind the evaluation is described in the section 6 of the supplementary material (\emph{Evaluation methodology}).

Briefly, for 68 categories from the FTC we can manually identify a set of semantically equivalent categories in the ATC. We call these equivalent categories the \emph{evaluation points} (see supplementary material section 6.1 for details). 
All drugs from each evaluation point were then assessed to determine the quality of the FTC against our gold standard, i.e.\ the ATC.
For example, the FTC category \emph{Anti-hydrogen:potassium-exchanging ATPase activity agent} (FTC\_A0008900) has 
been manually asserted as equivalent to the ATC category \emph{Proton pump inhibitors} (A02BC). A summary of this 
evaluation point is furthermore available online at {{https://www.ebi.ac.uk/chembl/ftc/evaluation/FTC\_A0008900}}.

For 1'280 DrugBank compounds in the FTC, 1'134 are also present in the ATC, therefore only those were 
considered. The \emph{evaluation points} cover a total of 471 DrugBank compounds or around 41\% of common drugs to both classifications.
Out of these, 275 compounds are true positives, i.e.\ they match both, the FTC and ATC categories for a given evaluation point.
The \emph{proton pump inhibitor} 
evaluation point is such a case where all the drugs (\emph{omeprazole, esomeprazole, pantoprazole, lansoprazole, rabeprazole}) are 
present both in the FTC category and in the corresponding ATC category. The total number of compounds from an ATC category but 
where we could not identify a corresponding FTC category is 35 (false negatives). Finally, 280 compounds are present in a FTC class but not in any 
corresponding ATC category (false positives). Overall we derive a 
recall of 89\%; this percentage indicates that the automatic build of the FTC covers a good portion of the content 
already present in the ATC. The precision of 50\% shows that the FTC contains for a given MoA many more drugs than the 
equivalent ATC categories. This result was expected and comes the original idea behind the FTC: Representing in a systematic fashion
the implicit and explicit MoAs of drugs, in particular the ones not already indexed by current classification scheme.

@@

\subsubsection{Polypharmacology spectrum}
The more information on a drug’s molecular targets and their physiological roles, the more opportunities 
exist to re-orient a drug into doing something new. The therapeutic agents described in the FTC can have several MoAs, 
i.e.\ may be acting on different biomolecular functions or processes, 
which demonstrates the intrinsic polypharmacology of the approved compounds. Figure \ref{fig:03} illustrates the polypharmacology 
spectrum by showing the distribution of number of MoAs per compound. When only direct categories are considered, compounds belong on average
to 13.5 MoA categories. This number increases to 61.2 when parent categories are taken into consideration (super classes).
Not all the MoAs are relevant to a disease, 
some FTC categories are particularly abstract (e.g, \emph{Anti-biological process agent}) yet they represent discrete 
categories to which the drug belongs with an explicit and clear meaning. These discrete MoAs are a good starting 
point to understand what a compound can do when administered in a human system.
Compound's polypharmacology is 
well represented in the FTC, as shown by the numerous MoAs each approved drug can exhibit.

We decided to further look at a well-known repositioning example, 
in order to see whether the FTC was suitable to identify the new
uses of an old drug. We picked the \emph{thalidomide} for 
this exercise ({{https://www.ebi.ac.uk/chembl/ftc/agent/DB01041}}). The molecule was 
first indicated to treat morning sickness
in pregnant women, but has been quickly abandoned after its developmental toxicity has been discovered in newborns. 
The accepted molecular mechanism behind the side-effect is  
an impairment of the angiogenic process responsible for the development of members, affecting in particular the limbs \citep{therapontos2009thalidomide}.
We found that the \emph{thalidomide} was accurately classified as \emph{Anti-cell migration involved in sprouting angiogenesis agent} in the FTC, capturing the known toxicity of the drug. Furthermore, the \emph{thalidomide} is currently investigated for a multitude of new usages, in particular for anti-cancer and immunomodulatory activities among others \citep{teo2005thalidomide}.
These new indications are well represented in the FTC too, for 
example by the categories \emph{Anti-vascular endothelial growth factor production agent} or 
\emph{Anti-cell division agent} for antineoplastic activities, 
or by the classes \emph{Anti-cytokine secretion agent} and \emph{Anti-I-kappaB kinase/NF-kappaB cascade agent} for its effect
on the immune system. These observations demonstrate that the FTC can successfully capture the 
molecular reasons behind the repositioning of an old compound. 
Moreover the classification can also provide valuable insight regarding
potential toxicity too.

\begin{figure*}[!tpb]%figure4
\centerline{\includegraphics{fig4.png}}
\caption{Pairwise comparison of MoAs similarities. Therapeutic indications are represented by ATC categories which are the colours on the side.
For instance, the compound \emph{reteplase} (DB00015) has the ATC code B01AD07, 
which appears as \emph{B} (dark orange) on the plot.
Only the first ATC level 
is considered. The reader can refer to Supplementary Figure 1A for a two ATC level granularity. 
The similarity descriptor ranges from 0 (not similar - white) to 1 (identical - black).
Some compounds belong to multiple ATC categories (\emph{Multiple}) and some others do not have
an ATC code (\emph{NoCategory}).
The average similarity of drugs present in the same therapeutic category 
is significantly higher on average when separately compared to all other indications.}\label{fig:04}
\end{figure*}

\subsubsection{Drugs with similar MoAs have similar indications}
The list of MoAs attributed to a drug can be exploited as a descriptor for the therapeutic agent:
The tree structure of the FTC can be used to derive some similarity metrics over the MoAs. The underlying 
heuristic is to assume that the closer two entities are in the taxonomy, the more similar they are. 
We used a straightforward approach derived from the Jaccard index (see supplementary section 7) in order to compare 
approved drugs based on the similarity of their MoAs. For instance, the similarity between two compounds present in 
the same FTC category is 1 (maximum). The similarity between an \emph{anti-blood coagulant} and \emph{pro-blood coagulant} is 0.29, 
reflecting the fact that such compounds are dissimilar with regards to the outcome of their biological effect. As the MoA is 
intuitively expected to be the central concept leading to the indication of the drug, we expected that on average, drugs with 
similar MoAs would be indicated towards similar therapeutic areas.

The heat map presented in Figure \ref{fig:04} shows a pairwise 
comparison of all the drugs of the FTC based on their relative MoA similarity. The compounds are further grouped by 
therapeutic indications as defined by the ATC.
The heatmap reveals some square patches around the central diagonal;
the overall similarity appears higher when compounds from the same ATC group are considered. 
A significance analysis (see supplement, section 8)
revealed that the average MoA similarity of compounds belonging to the same ATC category is 
significantly higher than when compounds belonging to different categories are compared. Indeed, for 
each category, the p-value was inferior to 0.05 based on 20'000 random permutations over the similarity values. 
This result supports the idea that drugs with similar MoAs have similar indications. Note that the mean of the 
similarity values was considered for the statistical analysis; some outliers are also present in the map, which 
can be interpreted as repositioning hypotheses. These outliers have indeed similar MoAs, yet they belong to totally 
different therapeutic areas and are used for different purposes according to the ATC. We are currently further analysing 
such cases in a systematic fashion. Hypotheses have to be manually examined and interpreted, as ATC categories are
only covering some of the legal usage of the drugs. We expect to find off-label indications in 
the predictions for instance, as well as false positives.

Supplementary Figure 3 present similar association behaviour when two levels of the ATC are 
considered (no statistical significance performed). Supplementary Figure 4 re-uses the same data as Figure \ref{fig:04} (one ATC level) 
but with a clustering function apply to it (hierarchical clustering - manhattan distance) in order to reveal 
functional clusters of drugs. Finally, Supplementary Figure 5 shows the distribution when compounds are sorted based on 
their identifiers; no patterns are identifiable in this case. Taken together, these results emphasise that the MoAs as 
defined in the FTC are indeed on average associated with the therapeutic indication of a drug. This result supports the validity of the
resource and its potential to computationally address indication discovery.
